Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
More preproc stuff

One candidate for such a predicate is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, exclusivity is not an \emph{inductive} invariant.
To see why, consider the following state <{s}>: let thread~0 be at label <{cs}>
and thread~1 be at the start of the \textbf{await} statement.
Also, in state <{s}>, $\mathit{turn} = 1$.  Now let
thread~1 make a step.  Because $\mathit{turn} = 1$,
thread~1 will stop waiting and also enter the critical
section, entering a state that is not exclusive.
So, exclusivity is an invariant (holds over every reachable state, as demonstrated
by Harmony),
but not an inductive invariant.
It will turn out that <{s}> is not reachable.

@@

We need a stronger inductive invariant than $\mathcal{F}(i)$ to prove mutual exclusion.
What else do we know when thread <{i}> is in the critical section?
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor \mathit{turn} = i$, that is,
the condition on the \textbf{await} statement for thread~<{i}>.
In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread <{i}> is in
the critical section: $\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i)$.
However, because thread $1-i$ is executing concurrently, this property does not
hold.
You can use Harmony to verify this.  Just place the following command
in the critical section of the program:

@@

The two things to note here is the use of the \textbf{choose}
expression and the \textbf{assert} statement.  Using <{choose}>,
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert}, we are able to express the anomaly we are
looking for.

@@

% In Harmony, atomic statements allow you to \emph{implement} your own
% atomic primitives such as test-and-set.  Other common examples
% include compare-and swap and fetch-and-add.
Atomic statements
are not intended to replace locks or other synchonization primitives.
When implementing synchronization solutions you should not directly use
atomic statements but use the synchronization primitives that are available
to you.  But if you want to \emph{specify} a synchronization primitive,
then use \textbf{atomically} by all means.
You can also use atomic statements in your test code.
In fact, as mentioned before, \textbf{assert} statements are included to
test if certain conditions hold in every execution and are executed
atomically.

@@

Harmony can check these invariants as well.
$\mathcal{I}_2(i)$ is checked by the \textbf{assert} statement.
But how would one go about checking an
invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
For $\mathcal{I}_2$ we only need an assertion at label <{cs}> because the
premise is that there is a thread at that label.  However, we would
like to check $\mathcal{I}_1$ in \emph{every state} (after the variables have
been initialized).
Harmony supports checking such invariants using the \texttt{invariant}
keyword.
The expression counts the number of <{False}> values and checks that
the result is less than or equal to 1.
Harmony checks the expression in every reachable state.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) <{synchS}>.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement <{tryAcquire(b)}> as an additional
interface for both the <{synch}> and <{synchS}> modules.
This interface is like <{acquire(b)}> but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

The reason is that \texttt{queue.get()} changes the state by
acquiring a lock, but the expressions in \textbf{assert}
statements (or \textbf{invariant} statements)
are not allowed to change the state.

@@

\begin{comment}
\autoref{fig:RWqueue} and \autoref{fig:RWqtest} present a different approach
in which threads are allowed into the critical section in a purely First
Come First Served manner.  When a reader arrives, it is allowed in only if
there is no writer in the critical section or waiting to enter.  When a writer
arrives, it is allowed in only if there is no reader or writer in the
critical section (and therefore also not waiting to enter).  To maintain the
arrival order, the solution uses a semaphore per thread and a queue.
The \texttt{acquire\_} methods take a pointer to
the semaphore of the thread as argument.
Each entry in the queue is a pair consisting of a type of access
(\texttt{.read} or \texttt{.write}) and a pointer to the semaphore of the
thread that is waiting.  Method \texttt{release\_one} considers semaphores
to release in the order in which they appear in the queue.

@@

\section*{Exercises}
\begin{problems}
\item Check if the code finds a unique leader if identifiers are not unique.
\item Messages are added atomically to the network.  Is this necessary?
What happens if you remove the \textbf{atomically} keyword?  Explain what happens.
\end{problems}

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a \texttt{PREPARE}
message to each of the servers involved in the transaction, asking them
to prepare to commit to their part in a particular transaction.
A server can either respond with
\texttt{YES} if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or with \texttt{NO} if it does not want to participate in the transaction.
If all servers respond with \texttt{YES}, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with \texttt{YES} (if any) must be notified
to inform them of the coordinator's decision.

@@

\noindent
Messages sent to a bank have the following dictionary format:
$\{ \mathtt{.dst}, \mathtt{.src}, \mathtt{.request}, \mathtt{.amount} \}$:
\begin{itemize}
\item \texttt{.dst}: identifier of the bank;
\item \texttt{.src}: identifier of the coordinator that sent the message;
\item \texttt{.request}: request type, which is either
\texttt{.withdraw}, \texttt{.deposit}, \texttt{.commit}, or \texttt{.abort};
\item \texttt{.amount}: amount to withdraw or deposit.
\end{itemize}

\noindent
A bank waits for a message destined to itself that it has not yet received.
In case of a withdrawal when the bank is idle and there are sufficient
funds, the bank saves the current balance in <{status}> to indicate
an ongoing transaction and what its original balance was.
The bank then responds with a \texttt{.yes} message to the coordinator,
including the new balance.
Otherwise, the bank responds with a \texttt{.no} message.
Deposits are similar, except that it is not necessary to check for
sufficient funds.
In case of a \texttt{.commit} message, the bank changes its status
to (), indicating that there is no ongoing transaction.  In case of
a \texttt{.abort} message, the bank restores <{balance}> first.

\autoref{fig:2pc2} contains the code for transfers and inquiries, as well
as tests.
The \texttt{receive()} method is used by coordinators in an
\textbf{atomically when exists}
statement to wait for a response from each bank involved in a transaction.
Argument \texttt{self} is the
identifier of the coordinator and \texttt{sources} is the set of banks.
It returns the empty set if there not yet responses from all banks.
Otherwise it returns a singleton set containing the set of responses,
one for each source.

The \texttt{transfer}() method contains the code for the transfer
transaction.  Argument <{self}> is the identifier of the coordinator,
<{b1}> is the source bank, <{b2}> is the destination bank,
and <{amt}> is the amount to transfer.
The coordinator sends a \texttt{PREPARE} message containing a
\texttt{.withdraw} request to <{b1}> and a \texttt{PREPARE}
message containing a \texttt{.deposit} request to <{b2}>.
It then waits for responses from each.  If both responses are
\texttt{.yes}, then it commits the transaction, otherwise it aborts
the transaction.

\begin{comment}
Note the use of the \textbf{possibly} statement in this code.
It is similar to \textbf{assert}, but where \texttt{assert} checks that
the predicate holds in \emph{all} executions, \texttt{possibly} checks
that the predicate holds in at least one execution.
The purpose is to check that the code doesn't simply always abort---there
exist executions where the code commits.
In practice, hopefully \emph{most} executions commit, but Harmony is not
capable of probabilistic reasoning.
\end{comment}

@@

Messages have the format $(\mathit{source}, \mathit{payload})$,
where <{source}> is the identifier of the replica that generated
the message and <{payload}> has information about the state of
the replica.  The payload can take one of two forms.  The first is
\texttt{.crash}, indicating that the source of the message has
crashed.  The code models a replica crashing by that replica broadcasting
a \texttt{.crash} message on the network.
The other form of payload is a tuple
$(\mathit{index}, \mathit{operation})$, which means that the source
placed the given operation at the given index in its history.

@@

Otherwise, the replica does not believe it is the head and
awaits a message from its direct predecessor.
If the payload is \texttt{.crash}, then the replica removes this
predecessor from its set of predecessors.  For example, if there
are initially three replicas numbered 0, 1, and 2, and replica~1
crashes, then replica~2 removes~1 from its set of predecessors, making
replica~0 its direct (and only) predecessor.  If, instead, the payload
is a tuple, then it must be the case that it contains the next operation
to add to its history.  The replica adds the operation to its history
and broadcasts the update, again intended for its successor.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a \texttt{.read}
request to all servers and then waiting for a response from
$\mathtt{N} - \mathtt{F}$ servers.
The \texttt{receive()} function uses the \texttt{bag.combinations} method
to find all combinations of subsets of responses of size
$\mathtt{N} - \mathtt{F}$.
The second phase of each operation is similar.

\autoref{fig:abdtest} tests the third property of the properties listed
above.  The \texttt{writer()} method writes a new value \texttt{.token}.
The \texttt{reader()} method reads twice and makes sure that if the
first read operation returns \texttt{.token}, then so does the second.
This illustrates the importance of the second phase of the \texttt{read}
operation.  You can comment out Lines~32 and~33 in \autoref{fig:abd} and
see what goes wrong.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{<{e}> == <{e}>} & equivalence \\
\hline
\texttt{<{e}> != <{e}>} & inequivalence \\
\hline
\texttt{<{e}> < <{e}>, <{e}> <= <{e}>, <{e}> > <{e}>, <{e}> >= <{e}>} & comparison\\
\hline
\end{tabular}
\end{center}

@@

Operations on booleans include:
\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{<{e}> and <{e}> and ...} & conjuction \\
\hline
\texttt{<{e}> or <{e}> or ...} & disjunction \\
\hline
\texttt{<{e}> => <{e}>, <{e}> not => <{e}>} & implication \\
\hline
\texttt{not <{e}>} & negation \\
\hline
\texttt{<{v}> if <{e}> else $v'$} & <{v}> or $v'$ depending on <{e}> \\
\hline
\texttt{any <{s}>, all <{s}>} & disjunction / conjunction for set or list <{s}> \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{-<{e}>} & negation \\
\hline
\texttt{abs(<{e}>)} & absolute value \\
\hline
\texttt{<{e}> + <{e}> + ...} & sum \\
\hline
\texttt{<{e}> - <{e}>} & difference \\
\hline
\texttt{<{e}> * <{e}> * <{e}> ...} & product \\
\hline
\texttt{<{e}> / <{e}>, <{e}> // <{e}>} & integer division \\
\hline
\texttt{<{e}> \% <{e}>, <{e}> mod <{e}>} & integer division remainder \\
\hline
\texttt{<{e}> ** <{e}>} & power \\
\hline
\texttt{\string~<{e}>} & binary inversion \\
\hline
\texttt{<{e}> \& <{e}> \& ...} & binary and \\
\hline
\texttt{<{e}> | <{e}> | ...} & binary or \\
\hline
\texttt{<{e}> \string^ <{e}> \string^ ...} & binary exclusive or \\
\hline
\texttt{$e << e$} & binary shift left \\
\hline
\texttt{$e >> e$} & binary shift right \\
\hline
\texttt{$\{ e..e' \}$} & set of integers from <{e}> to $e'$ inclusive \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{len <{s}>} & cardinality \\
\hline
\texttt{<{s}> - <{s}>} & set difference \\
\hline
\texttt{<{s}> \& <{s}> \& ...} & intersection \\
\hline
\texttt{<{s}> | <{s}> | ...} & union \\
\hline
\texttt{<{s}> \string^ <{s}> \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
\hline
\texttt{choose <{s}>} & select an element (Harmony will try all) \\
\hline
\texttt{min <{s}>} & minimum element \\
\hline
\texttt{max <{s}>} & maximum element \\
\hline
\texttt{any <{s}>} & <{True}> if any value is <{True}> \\
\hline
\texttt{all <{s}>} & <{True}> if all values are <{True}> \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$d~k$} & indexing \\
\hline
\texttt{len <{d}>} & the number of keys in <{d}> \\
\hline
\texttt{keys <{d}>} & the set of keys in <{d}> \\
\hline
\texttt{<{v}> [not] in keys <{d}>} & check if <{v}> is a \emph{key} in <{d}> \\
\hline
\texttt{<{v}> [not] in <{d}>} & check if <{v}> is a \emph{value} in <{d}> \\
\hline
\texttt{min <{d}>} & the minimum value in <{d}> \\
\hline
\texttt{max <{d}>} & the maximum value in <{d}> \\
\hline
\texttt{any <{d}>} & <{True}> if any value is <{True}> \\
\hline
\texttt{all <{d}>} & <{True}> if all values are <{True}> \\
\hline
\texttt{<{d}> \& <{d}> \& ...} & dictionary intersection \\
\hline
\texttt{<{d}> | <{d}> | ...} & dictionary union \\
\hline
\end{tabular}
\end{center}

Because in Harmony brackets are used for parsing purposes only,
you can write $d[k]$ (or $d(k)$) instead of $d~k$.
However, if <{k}> is a string like \texttt{.id}, then you might
prefer the notation $k\mathtt{.id}$.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$t~k$} & indexing \\
\hline
\texttt{<{t}> + <{t}> + ...} & concatenation \\
\hline
\texttt{<{t}> * $n$} & $n$ copies of <{t}> concatenated \\
\hline
\texttt{<{v}> [not] in <{t}>} & check if <{v}> is a \emph{value} in <{t}> \\
\hline
\texttt{len <{t}>} & the length of <{t}> \\
\hline
\texttt{min <{t}>} & the minimum value in <{t}> \\
\hline
\texttt{max <{t}>} & the maximum value in <{t}> \\
\hline
\texttt{any <{t}>} & <{True}> if any value is <{True}> \\
\hline
\texttt{all <{t}>} & <{True}> if all values are <{True}> \\
\hline
\texttt{<{t}> \& <{t}> \& ...} & pairwise list minimum \\
\hline
\texttt{<{t}> | <{t}> | ...} & pairwise list maximum \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{<{v}> [not] in keys <{b}>} & check if <{v}> is in <{b}> \\
\hline
\texttt{<{t}> \& <{t}> \& ...} & bag intersection\\
\hline
\texttt{<{t}> | <{t}> | ...} & bag union \\
\hline
\end{tabular}
\end{center}

@@

In Harmony, each shared variable has an address, which is essentially
a list of Harmony values.
For example, the address of variable $d.\mathtt{v}[3]$ consists of the
list \texttt{.d}, \texttt{.v}, and 3.  The only way to construct an
address in Harmony is using the $?$ operator.
$?d.\mathtt{v}[3]$ returns the address of variable $d.\mathtt{v}[3]$.
An address can be dereferenced to return the value of the variable.
If <{a}> is an address, then $!a$ returns the value.
Like C, Harmony supports the shorthand \texttt{<{a}>->v} for 
the expression \texttt{($!a$).v}.

@@

\subsection*{\textbf{assert}}

The statement \textbf{assert <{b}>} evaluates <{b}> and reports an error
if <{b}> is false.  It should be considered a no-op---it is part of the
\emph{specification}, not part of the \emph{implementation} of the algorithm.
In particular, it specifies an invariant: whenever the program counter is
at the location where the \textbf{assert} statement is, then <{b}> is
always true.

@@

As in Python, you can specify an additional expression:
\textbf{assert <{b}>, <{e}>}.  The value of <{e}> will be reported as part
of the error should <{b}> evaluate to false.

\subsection*{\textbf{atomically}}

@@

\textbf{atomically} statements are useful for specification and
implementing synchronization primitives such as test-and-set.
It is also useful for testing.
It is not a replacement for lock/unlock, and should not generally be used
for synchronization otherwise.  Lock/unlock does allow other
threads to run concurrently---just not in the same critical section.

\subsection*{\textbf{await}}

The statement \textbf{await <{b}>} is equivalent to \textbf{when <{b}>: pass}.
It is intended to improve readability of your code.

\subsection*{\textbf{const}}

@@

\subsection*{\textbf{def}}

@@

\subsection*{\textbf{del}}

The statement \textbf{del} <{x}> removes variable <{x}> from the
state.  <{x}> can be either a local or a shared variable.  For
example, the statement \textbf{del} $x.\texttt{age}$ removes
the \texttt{.age} field from dictionary <{x}>.
Harmony automatically removes top-level local variables that
are no longer in use
from the state in order to attempt to reduce the number of states
that are evaluated during model checking.

>>>>

One candidate for such a predicate is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, exclusivity is not an \emph{inductive} invariant.
To see why, consider the following state <{s}>: let thread~0 be at label <{cs}>
and thread~1 be at the start of the <{await}> statement.
Also, in state <{s}>, $\mathit{turn} = 1$.  Now let
thread~1 make a step.  Because $\mathit{turn} = 1$,
thread~1 will stop waiting and also enter the critical
section, entering a state that is not exclusive.
So, exclusivity is an invariant (holds over every reachable state, as demonstrated
by Harmony),
but not an inductive invariant.
It will turn out that <{s}> is not reachable.

@@

We need a stronger inductive invariant than $\mathcal{F}(i)$ to prove mutual exclusion.
What else do we know when thread <{i}> is in the critical section?
Let $\mathcal{C}(i) = \lnot\mathit{flags}[1 - i] \lor \mathit{turn} = i$, that is,
the condition on the <{await}> statement for thread~<{i}>.
In a sequential program, $\mathcal{C}(i)$ would clearly hold if thread <{i}> is in
the critical section: $\mathtt{thread}(i)@cs \Rightarrow \mathcal{C}(i)$.
However, because thread $1-i$ is executing concurrently, this property does not
hold.
You can use Harmony to verify this.  Just place the following command
in the critical section of the program:

@@

The two things to note here is the use of the \textbf{choose}
expression and the <{assert}> statement.  Using <{choose}>,
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert}, we are able to express the anomaly we are
looking for.

@@

% In Harmony, atomic statements allow you to \emph{implement} your own
% atomic primitives such as test-and-set.  Other common examples
% include compare-and swap and fetch-and-add.
Atomic statements
are not intended to replace locks or other synchonization primitives.
When implementing synchronization solutions you should not directly use
atomic statements but use the synchronization primitives that are available
to you.  But if you want to \emph{specify} a synchronization primitive,
then use <{atomically}> by all means.
You can also use atomic statements in your test code.
In fact, as mentioned before, <{assert}> statements are included to
test if certain conditions hold in every execution and are executed
atomically.

@@

Harmony can check these invariants as well.
$\mathcal{I}_2(i)$ is checked by the <{assert}> statement.
But how would one go about checking an
invariant like $\mathcal{I}_1$?  Invariants must hold for every state.
For $\mathcal{I}_2$ we only need an assertion at label <{cs}> because the
premise is that there is a thread at that label.  However, we would
like to check $\mathcal{I}_1$ in \emph{every state} (after the variables have
been initialized).
Harmony supports checking such invariants using the \texttt{invariant}
keyword.
The expression counts the number of <{False}> values and checks that
the result is less than or equal to 1.
Harmony checks the expression in every reachable state.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) <{synchS}>.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement <{tryAcquire(b)}> as an additional
interface for both the <{synch}> and <{synchS}> modules.
This interface is like <{acquire(b)}> but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the <{atomically}>
keyword.
\end{problems}

@@

The reason is that \texttt{queue.get()} changes the state by
acquiring a lock, but the expressions in <{assert}>
statements (or \textbf{invariant} statements)
are not allowed to change the state.

@@

\begin{comment}
\autoref{fig:RWqueue} and \autoref{fig:RWqtest} present a different approach
in which threads are allowed into the critical section in a purely First
Come First Served manner.  When a reader arrives, it is allowed in only if
there is no writer in the critical section or waiting to enter.  When a writer
arrives, it is allowed in only if there is no reader or writer in the
critical section (and therefore also not waiting to enter).  To maintain the
arrival order, the solution uses a semaphore per thread and a queue.
The \texttt{acquire\_} methods take a pointer to
the semaphore of the thread as argument.
Each entry in the queue is a pair consisting of a type of access
(<{.read}> or <{.write}>) and a pointer to the semaphore of the
thread that is waiting.  Method \texttt{release\_one} considers semaphores
to release in the order in which they appear in the queue.

@@

\section*{Exercises}
\begin{problems}
\item Check if the code finds a unique leader if identifiers are not unique.
\item Messages are added atomically to the network.  Is this necessary?
What happens if you remove the <{atomically}> keyword?  Explain what happens.
\end{problems}

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a <{PREPARE}>
message to each of the servers involved in the transaction, asking them
to prepare to commit to their part in a particular transaction.
A server can either respond with
\texttt{YES} if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or with \texttt{NO} if it does not want to participate in the transaction.
If all servers respond with \texttt{YES}, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with \texttt{YES} (if any) must be notified
to inform them of the coordinator's decision.

@@

\noindent
Messages sent to a bank are dictionaries with the following fields:
\begin{itemize}
\item <{.dst}>: identifier of the bank;
\item <{.src}>: identifier of the coordinator that sent the message;
\item <{.request}>: request type, which is either
<{.withdraw}>, <{.deposit}>, <{.commit}>, or <{.abort}>;
\item <{.amount}>: amount to withdraw or deposit.
\end{itemize}

\noindent
A bank waits for a message destined to itself that it has not yet received.
In case of a withdrawal when the bank is idle and there are sufficient
funds, the bank saves the current balance in <{status}> to indicate
an ongoing transaction and what its original balance was.
The bank then responds with a <{.yes}> message to the coordinator,
including the new balance.
Otherwise, the bank responds with a <{.no}> message.
Deposits are similar, except that it is not necessary to check for
sufficient funds.
In case of a <{.commit}> message, the bank changes its status
to (), indicating that there is no ongoing transaction.  In case of
a <{.abort}> message, the bank restores <{balance}> first.

\autoref{fig:2pc2} contains the code for transfers and inquiries, as well
as tests.
The <{receive()}> method is used by coordinators in an
\textbf{atomically when exists}
statement to wait for a response from each bank involved in a transaction.
Argument \texttt{self} is the
identifier of the coordinator and \texttt{sources} is the set of banks.
It returns the empty set if there not yet responses from all banks.
Otherwise it returns a singleton set containing the set of responses,
one for each source.

The \texttt{transfer}() method contains the code for the transfer
transaction.  Argument <{self}> is the identifier of the coordinator,
<{b1}> is the source bank, <{b2}> is the destination bank,
and <{amt}> is the amount to transfer.
The coordinator sends a <{PREPARE}> message containing a
<{.withdraw}> request to <{b1}> and a <{PREPARE}>
message containing a <{.deposit}> request to <{b2}>.
It then waits for responses from each.  If both responses are
<{.yes}>, then it commits the transaction, otherwise it aborts
the transaction.

\begin{comment}
Note the use of the \textbf{possibly} statement in this code.
It is similar to <{assert}>, but where \texttt{assert} checks that
the predicate holds in \emph{all} executions, \texttt{possibly} checks
that the predicate holds in at least one execution.
The purpose is to check that the code doesn't simply always abort---there
exist executions where the code commits.
In practice, hopefully \emph{most} executions commit, but Harmony is not
capable of probabilistic reasoning.
\end{comment}

@@

Messages have the format $(\mathit{source}, \mathit{payload})$,
where <{source}> is the identifier of the replica that generated
the message and <{payload}> has information about the state of
the replica.  The payload can take one of two forms.  The first is
<{.crash}>, indicating that the source of the message has
crashed.  The code models a replica crashing by that replica broadcasting
a <{.crash}> message on the network.
The other form of payload is a tuple
$(\mathit{index}, \mathit{operation})$, which means that the source
placed the given operation at the given index in its history.

@@

Otherwise, the replica does not believe it is the head and
awaits a message from its direct predecessor.
If the payload is <{.crash}>, then the replica removes this
predecessor from its set of predecessors.  For example, if there
are initially three replicas numbered 0, 1, and 2, and replica~1
crashes, then replica~2 removes~1 from its set of predecessors, making
replica~0 its direct (and only) predecessor.  If, instead, the payload
is a tuple, then it must be the case that it contains the next operation
to add to its history.  The replica adds the operation to its history
and broadcasts the update, again intended for its successor.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a <{.read}>
request to all servers and then waiting for a response from
<{N - F}> servers.
The <{receive()}> function uses the <{bag.combinations}> method
to find all combinations of subsets of responses of size
$\mathtt{N} - \mathtt{F}$.
The second phase of each operation is similar.

\autoref{fig:abdtest} tests the third property of the properties listed
above.  The <{writer()}> method writes a new value <{.token}>.
The \texttt{reader()} method reads twice and makes sure that if the
first read operation returns <{.token}>, then so does the second.
This illustrates the importance of the second phase of the \texttt{read}
operation.  You can comment out Lines~32 and~33 in \autoref{fig:abd} and
see what goes wrong.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{e == e}> & equivalence \\
\hline
<{e != e}> & inequivalence \\
\hline
<{e < e, e <= e, e > e, e >= e}> & comparison\\
\hline
\end{tabular}
\end{center}

@@

Operations on booleans include:
\begin{center}
\begin{tabular}{|l|l|}
\hline
<{e and e and ...}> & conjuction \\
\hline
<{e or e or ...}> & disjunction \\
\hline
<{e => e, e not => e}> & implication \\
\hline
<{not e}> & negation \\
\hline
<{v if e else v'}> & <{v}> or <{v'}> depending on <{e}> \\
\hline
<{any s, all s}> & disjunction / conjunction for set or list <{s}> \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{-e}> & negation \\
\hline
<{abs(e)}> & absolute value \\
\hline
<{e + e + ...}> & sum \\
\hline
<{e - e}> & difference \\
\hline
<{e * e * e ...}> & product \\
\hline
<{e / e, e // e}> & integer division \\
\hline
<{e % e, e mod e}> & integer division remainder \\
\hline
<{e ** e}> & power \\
\hline
<{~e}> & binary inversion \\
\hline
<{e & e & ...}> & binary and \\
\hline
<{e | e | ...}> & binary or \\
\hline
<{e ^ e ^ ...}> & binary exclusive or \\
\hline
<{e << e}> & binary shift left \\
\hline
<{e >> e}> & binary shift right \\
\hline
<{{ e..e' }}> & set of integers from <{e}> to $e'$ inclusive \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{len s}> & cardinality \\
\hline
<{s - s}> & set difference \\
\hline
<{s & s & ...}> & intersection \\
\hline
<{s | s | ...}> & union \\
\hline
<{s ^ s ^ ...}> & inclusion/exclusion (elements in odd number of sets) \\
\hline
<{choose s}> & select an element (Harmony will try all) \\
\hline
<{min s}> & minimum element \\
\hline
<{max s}> & maximum element \\
\hline
<{any s}> & <{True}> if any value is <{True}> \\
\hline
<{all s}> & <{True}> if all values are <{True}> \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{d k}> & indexing \\
\hline
<{len d}> & the number of keys in <{d}> \\
\hline
<{keys d}> & the set of keys in <{d}> \\
\hline
<{<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
\hline
<{v [not] in d}> & check if <{v}> is [not] a value in <{d}> \\
\hline
<{min d}> & the minimum value in <{d}> \\
\hline
<{max d}> & the maximum value in <{d}> \\
\hline
<{any d}> & <{True}> if any value is <{True}> \\
\hline
<{all d}> & <{True}> if all values are <{True}> \\
\hline
<{d & d & ...}> & dictionary intersection \\
\hline
<{d | d | ...}> & dictionary union \\
\hline
\end{tabular}
\end{center}

Because in Harmony brackets are used for parsing purposes only,
you can write $d[k]$ (or $d(k)$) instead of $d~k$.
However, if <{k}> is a string like <{.id}>, then you might
prefer the notation $k\mathtt{.id}$.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{t k}> & indexing \\
\hline
<{t + t + ...}> & concatenation \\
\hline
<{t * n}> & <{n}> copies of <{t}> concatenated \\
\hline
<{v [not] in t}> & check if <{v}> is [not] a value in <{t}> \\
\hline
<{len t}> & the length of <{t}> \\
\hline
<{min t}> & the minimum value in <{t}> \\
\hline
<{max t}> & the maximum value in <{t}> \\
\hline
<{any t}> & <{True}> if any value is <{True}> \\
\hline
<{all t}> & <{True}> if all values are <{True}> \\
\hline
<{t & t & ...}> & pairwise list minimum \\
\hline
<{t | t | ...}> & pairwise list maximum \\
\hline
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{v [not] in keys b}> & check if <{v}> is [not] in <{b}> \\
\hline
<{t & t & ...}> & bag intersection\\
\hline
<{t | t | ...}> & bag union \\
\hline
\end{tabular}
\end{center}

@@

In Harmony, each shared variable has an address, which is essentially
a list of Harmony values.
For example, the address of variable <{d.v[3]}> consists of the
list <{.d}>, <{.v}>, and 3.  The only way to construct an
address in Harmony is using the <{?}> operator.
<{?d.v[3]}> returns the address of variable <{d.v[3]}>.
An address can be dereferenced to return the value of the variable.
If <{a}> is an address, then <{!a}> returns the value.
Like C, Harmony supports the shorthand <{a->v}> for the expression <{(!a).v}>.

@@

\subsection*{<{assert}>}

The statement <{assert b}> evaluates <{b}> and reports an error
if <{b}> is false.  It should be considered a no-op---it is part of the
\emph{specification}, not part of the \emph{implementation} of the algorithm.
In particular, it specifies an invariant: whenever the program counter is
at the location where the <{assert}> statement is, then <{b}> is
always true.

@@

As in Python, you can specify an additional expression:
<{assert b, e}>.  The value of <{e}> will be reported as part
of the error should <{b}> evaluate to false.

\subsection*{<{atomically}>}

@@

<{atomically}> statements are useful for specification and
implementing synchronization primitives such as test-and-set.
It is also useful for testing.
It is not a replacement for lock/unlock, and should not generally be used
for synchronization otherwise.  Lock/unlock does allow other
threads to run concurrently---just not in the same critical section.

\subsection*{<{await}>}

The statement <{await b}> is equivalent to \textbf{when <{b}>: pass}.
It is intended to improve readability of your code.

\subsection*{<{const}>}

@@

\subsection*{<{def}>}

@@

\subsection*{<{del}>}

The statement <{del x}> removes variable <{x}> from the
state.  <{x}> can be either a local or a shared variable.  For
example, the statement <{del x.age}> removes
the <{.age}> field from dictionary <{x}>.
Harmony automatically removes top-level local variables that
are no longer in use
from the state in order to attempt to reduce the number of states
that are evaluated during model checking.

Because Harmony lists are dictionaries, deleting from lists is different
from Python:
<{x = [.a, .b]; del x[0]}> results in <{x}> having value
<{{1: .b}}> rather than <{[.b]}> (which is <{{0: .b}}>).
